Implicit Neural Representation for Potential Field Geophysics

### 2023-10-11 discussion.tex
Update per DW comments, part 2

When a domain expert encounters low-resolution grid data, they may observe which gridding method was used, the line spacing and direction, and form their understanding of potential aliasing and spurious anomalies in the grid accordingly.
Assisted by this understanding, they can infer geological features with varying degrees of confidence.
The SR method however does not consider these constraints, because imputation is purely based on features extractable by the model.
This was demonstrated to introduce noise or artefacts during SR inference, as well as the incorrect enhancement of artifacts in low-resolution data such as aliased geological strike.
However, given the uncertainty of geological interpretation of geophysics data \parencite{sivarajahIdentifyingEffectiveInterpretation2013} SR outputs may be used as an auxiliary data when interpreting low-resolution magnetic features.
The domain expert can then use the SR grid to reassure or reassess the original low-resolution grid interpretation and improve their confidence accordingly.

Super-resolution for geophysics is still in its infancy and requires further efforts towards improving reliability before wide-spread adoption in practice.
This section proposes four avenues for future research:
\begin{enumerate}
    \item{} Exploring the outcomes from the use of SR in the task of geological mapping
    \item{} Understanding the impact of choices in constructing a model and dataset for SR geophysics for a range of geological domains,
    \item{} Identifying an improved image quality assessment for gridded potential field data.
    \item{} Investigating the use of a physics-informed approach for SR and implicit representation, and,
\end{enumerate}

\section{SR impact on interpretation}
The super-resolution outputs from the methods presented in this thesis have improved structural accuracy over the low-resolution grids.
This is likely to have a positive impact on the detection of features such as lineaments and unit boundaries for geological interpretation from magnetic data.
However, despite the reduction of introduced structural inaccuracies with the use of open access training data in \Cref{ch:paper2}, it is likely that some features with specific structural geometries will still be incorrectly super-resolved.
A domain expert may be able to recognise structures that are inconsistent using auxiliary information and the regional geological context, but machine processed super-resolution grids will use these data as provided.
In order to improve the adoption of SR for geological interpretation from potential field data, a rigorous study of the impact of SR on interpretation outcomes is required.
This could be a geological mapping task, such as the reinterpretation of regional geology using super-resolution upscaled regional magnetic TMI grids.
Comparison could then be made for the accuracy and benefit of a super-resolution method compared to the existing low-resolution interpretation.
However, such analysis will be difficult to perform with respect to different interpreters' variability in pattern recognition, especially in the context of their confidence in the reliability of the super-resolution product.
Previously, \textcite{sivarajahIdentifyingEffectiveInterpretation2013} reported significant variability between individuals when detecting geological targets from magnetic geophysics data.
A more reliable, repeatable and quantifiable approach would use automated pattern recognition methods developed for geophysical features.
For example, the automated lineament detection technique of \textcite{holdenAutomaticIdentificationResponses2011,holdenAutomatedAnalysisRegional2008} could be applied to low-resolution and SR grids, to compare their delineated features.

\section{SR model construction}
Several models were implemented throughout \Cref{ch:paper1,ch:paper2}, namely RDN, ESRGAN+, and LTE\@.
% However, the most important choice in developing a performant super-resolution neural network for real-world aeromagnetic survey data was the choice of training data.
\Cref{ch:paper1} was prepared prior the release of a large and varied synthetic geology training dataset \parencite{jessellNoddyverseMassiveData2022}.
It was seen in \Cref{ch:paper2} that this data improved model performance for a range of features by using transfer learning for pre-training.
% This transfer learning approach for training data is also found to be useful for RDN and ESRGAN\@.
Here, the use of synthetic and real-world line spacing sampled data approach used with LTE in \Cref{ch:paper2} are adopted for the RDN+ and ESRGAN+ models.
\Cref{fig:modcomp} shows RDN and ESRGAN+ in predicting accurate features when upscaling \qty{320}{\m} data to \qty{80}{\m} line spacing.
It can be inferred that the challenge of super-resolution for these tasks is predominantly a question of providing suitable and sufficient training data, and is less constrained to neural network choices.

@@

Magnetic data in different geological contexts are subject to a mixture of statistical distributions, \parencite{khokhlovCauseNonGaussianDistribution2017}. Investigation is also required to identify a robust normalisation method accounting for dynamic range and the structural significance contained in the long tails of the distributions of magnetic intensities.

@@

\section{Physics-informed neural networks}
Potential fields are constrained by several well understood physical laws, and gridded data ideally conform to these properties for use in inversion applications and further numerical processing.
Geological interpretation is less reliant on this adherence, however \emph{a priori} information used in physics-informed neural networks improve training speed and model performance \parencite{raissiPhysicsinformedNeuralNetworks2019}, and are especially useful in data-constrained domains such as geophysics.
It is possible to broaden the definition and extend the concept of informed networks to the use of generative adversarial networks in \Cref{ch:paper1}.
In these networks, the informing method is termed discriminator loss, and is performed by a classification network conditioned to identify images as `realistic' or `not realistic', depending on features learnt to be important within the target data by the discriminator.
Using crafted or learnt criterions to inform a neural network greatly assists with learning high quality models, and either approach may be useful for to geophysics in addressing the paucity of ground truth.
In addition, demonstrating the enforcement of physical properties of potential fields for model outputs may address apprehension for machine learning in geophysics practice.

@@

To the best of my knowledge, \Cref{ch:paper1} presents the first investigation in the literature of super-resolution enhancement of gridded aerogeophysical data.
Prior to this work, it was unknown how well SR methods developed for natural images in computer vision research would adapt to the features present within magnetic grids, or the challenges faced in developing a training dataset for these data.
We established that the CNN SR methods RDN and ESRGAN could be trained to upsample low-resolution textures in magnetic grids by four times scale, and that the results were more accurate than bicubic upsampling in the real-world case study region.
The GAN method used in \Cref{ch:paper2} enhanced resolution, however it introduced significant spurious features, masking the enhancement.
Hence, the method based on RDN was preferred.
We identified the need for sufficient representative training data, and a realistic method to generate low-resolution training data.

@@

The contribution in \Cref{ch:paper3} used coordinate multilayer perceptron neural networks to learn a representation of the function that describes a surveyed potential field extent.
To my knowledge, this is the first time implicit neural representation has been reported for use with aerogeophysical survey data in the literature.
These networks are highly suited for processing spatial data such as point sampled geophysics surveys.
The contribution is capable of predicting regularly spaced grids from scattered survey point data, and a novel method using the representation to calculate derivatives of a potential field, both fundamental to interpretation and processing in geophysics.
By commencing the methods from point sample data, the number of individual processing steps is reduced, and the application is fully trainable.
Coordinate MLPs for implicit representation are a currently burgeoning field of research in computer science, and we identify opportunities in extending advances in this research to future applications in geophysics, as outlined in \Cref{ch:futurework}.

>>>>

When a domain expert encounters low-resolution grid data, they may observe which gridding method was used, the line spacing and direction, and form their understanding of potential aliasing and spurious anomalies in the grid accordingly.
Assisted by this understanding, they can infer geological features with varying degrees of confidence.
The SR method however does not consider these constraints, because imputation is purely based on features extractable by the model.
This was demonstrated to introduce noise or artefacts during SR inference, as well as the incorrect enhancement of artefacts in low-resolution data such as aliased geological strike.
However, given the uncertainty of geological interpretation of geophysics data \parencite{sivarajahIdentifyingEffectiveInterpretation2013} SR outputs may be used as an auxiliary data when interpreting low-resolution magnetic features.
The domain expert can then use the SR grid to reassure or reassess the original low-resolution grid interpretation and improve their confidence accordingly.

Super-resolution for geophysics is still in its infancy and requires further efforts towards improving reliability before wide-spread adoption in practice.
This section proposes four avenues for future research:
\begin{enumerate}
    \item{} Exploring the impact on geological mapping from the use of SR,
    \item{} Understanding the impact of choices in training super-resolution for domain specific tasks,
    \item{} Identifying an improved image quality assessment for gridded potential field data and,
    \item{} Investigating the use of a physics-informed approach for SR and implicit representation.
\end{enumerate}

\section[Super-resolution impact on geological interpretation]{Super-resolution impact on geological interpretation}
The super-resolution outputs from the methods presented in this thesis have improved structural accuracy over the low-resolution grids.
This is likely to have a positive impact on the detection of features such as lineaments and unit boundaries for geological interpretation from magnetic data.
However, despite the reduction of introduced structural inaccuracies with the use of open access training data in \Cref{ch:paper2}, it is likely that some features with specific structural geometries will still be incorrectly super-resolved.
A domain expert may be able to recognise structures that are inconsistent using auxiliary information and the regional geological context, but machine processed super-resolution grids will use these data as provided.
In order to improve the adoption of SR for geological interpretation from potential field data, a rigorous study of the impact of SR on interpretation outcomes is required.
This could be a geological mapping task, such as the reinterpretation of regional geology using super-resolution upscaled regional magnetic TMI grids.
Comparison could then be made for the accuracy and benefit of a super-resolution method compared to the existing low-resolution interpretation.
However, such analysis will be difficult to perform with respect to different interpreters' variability in pattern recognition, especially in the context of their confidence in the reliability of the super-resolution product.
Previously, \textcite{sivarajahIdentifyingEffectiveInterpretation2013} reported significant variability between individuals when detecting geological targets from magnetic geophysics data.
A more reliable, repeatable and quantifiable approach would use automated pattern recognition methods developed for geophysical features.
For example, the automated lineament detection technique of \textcite{holdenAutomaticIdentificationResponses2011,holdenAutomatedAnalysisRegional2008} could be applied to low-resolution and SR grids, to compare their delineated features.

\section[Super-resolution for domain specific tasks]{A structured approach to training super-resolution for domain specific tasks}
Several models were implemented throughout \Cref{ch:paper1,ch:paper2}, namely RDN, ESRGAN+, and LTE\@.
% However, the most important choice in developing a performant super-resolution neural network for real-world aeromagnetic survey data was the choice of training data.
The work in \Cref{ch:paper1} was prepared for publication prior the release of a large and varied synthetic geology training dataset \parencite{jessellNoddyverseMassiveData2022}.
It was seen in \Cref{ch:paper2} that this data improved model performance for a range of features by using transfer learning for pre-training.
% This transfer learning approach for training data is also found to be useful for RDN and ESRGAN\@.
Here, the use of synthetic and real-world line spacing sampled data approach used with LTE in \Cref{ch:paper2} are adopted for the RDN+ and ESRGAN+ models.
\Cref{fig:modcomp} shows RDN and ESRGAN+ in predicting accurate features when upscaling \qty{320}{\m} data to \qty{80}{\m} line spacing.
It can be inferred that the challenge of super-resolution for these tasks is predominantly a question of providing suitable and sufficient training data, and is less constrained to neural network choices.

@@

Magnetic data in different geological contexts are subject to a mixture of statistical distributions \parencite{khokhlovCauseNonGaussianDistribution2017}. Investigation is also required to identify a robust normalisation method accounting for dynamic range and the structural significance contained in the long tails of the distributions of magnetic intensities.

@@

\section{Geophysics-informed neural networks}
Potential fields are constrained by several well understood physical laws, and gridded data ideally conform to these properties for use in inversion applications and further numerical processing.
Geological interpretation is less reliant on this adherence, however \emph{a priori} information used in physics-informed neural networks improve training speed and model performance \parencite{raissiPhysicsinformedNeuralNetworks2019}, and are especially useful in data-constrained domains such as geophysics.
It is possible to broaden the definition and extend the concept of informed networks to the use of generative adversarial networks in \Cref{ch:paper1}.
In these networks, the informing method is termed discriminator loss, and is performed by a classification network conditioned to identify images as `realistic' or `not realistic', depending on features learnt to be important within the target data by the discriminator.
Using crafted or learnt criterions to inform a neural network greatly assists with learning high quality models, and either approach may be useful for to geophysics in addressing the paucity of ground truth.
In addition, demonstrating the enforcement of physical properties of potential fields for model outputs may address apprehension for machine learning in geophysics practice.

@@

To the best of my knowledge, \Cref{ch:paper1} presents the first investigation in the literature of the super-resolution enhancement of gridded aerogeophysical data.
Prior to this work, it was unknown how well SR methods developed for natural images in computer vision research would adapt to the features present within magnetic grids, or the challenges faced in developing a training dataset for these data.
We established that the CNN SR methods RDN and ESRGAN could be trained to upsample low-resolution textures in magnetic grids by four times scale, and that the results were more accurate than bicubic upsampling in the real-world case study region.
The GAN method used in \Cref{ch:paper2} enhanced resolution, however it introduced significant spurious features, masking the enhancement.
Hence, the method based on RDN was preferred.
We identified the need for additional representative training data to investigate transferability of the method to other geological domains, and a realistic method to generate low-resolution training data.

@@

The contribution in \Cref{ch:paper3} used coordinate multilayer perceptron neural networks to learn a representation of the function that describes a surveyed potential field extent.
To my knowledge, this is the first time implicit neural representation of aerogeophysical survey data has been reported in the literature.
These networks are highly suited for processing spatial data such as point sampled geophysics surveys.
The contribution is capable of predicting regularly spaced grids from scattered survey point data, and a novel method using the representation to calculate derivatives of a potential field, both fundamental to interpretation and processing in geophysics.
By commencing the methods from point sample data, the number of individual processing steps is reduced, and the application is fully trainable.
Continued advances in the recent field of Coordinate MLPs will improve the capacity of the method for representing high-resolution aeromagnetic data. 
We identify opportunities in \Cref{ch:futurework} for future work in super-resolution and implicit representation for enhancing potential field methods in geophysics.
